Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-10-16 manuscript/manuscript_file.Rmd
Minor manuscript updates

Lakes, ponds, and reservoirs are thought to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020] have called into question the accuracy of emission estimates[@Lauerwald_etal_2019]. Here we measured dissolved N~2~O concentrations and calculated emission rates across `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies in the continental U.S. (CONUS). Nitrous oxide undersaturation was widespread through the CONUS, with `r propN2oSinkMean` % of the waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, such that all ecoregions functioned as net N~2~O sources to the atmosphere. Waterbodies with high nitrate (NO~3~) tended to be sources of N~2~O whereas those with low NO~3~ were generally sinks. Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. We estimate that CONUS waterbodies emit `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` ) metric tons N~2~O year^-1^, equivalent to only `r (national_stats %>% select(UCL_fn2o) %>% pull() / 22767) * 100`% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates . The demonstrated capacity for widespread N~2~O uptake in CONUS waterbodies highlights the need to revisit global N~2~O models which presume surface waters are a N~2~O source [@epa_2023].

>>>>

Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020] have called the accuracy of emission estimates into question [@Lauerwald_etal_2019]. Here we measured dissolved N~2~O concentrations and calculated emission rates for `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies distributed across the nine major ecoregions of the conterminous U.S. (CONUS). We use a Bayesian hierarchical approach to model the data and predicted N~2~O in`r all_predictions %>% distinct(.row) %>% summarize(nlake = n()) %>% pull()` CONUS waterbodies.  We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r propN2oSinkMean` % of the waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, and none of the ecoregions are estimated to functione as net N~2~O sinks. The model predicts that waterbodies with high nitrate (NO~3~) tend to be sources of N~2~O whereas those with low NO~3~ are generally sinks. Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS waterbodies is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` ) metric tons N~2~O year^-1^. The mean estimate is a net emission of  `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^, which is small relative to other known sources of N2O in the CONUS (e.g., wastewater treatment, manure management) and equivalent to only `r (national_stats %>% select(UCL_fn2o) %>% pull() / 22767) * 100`% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates . Emission estimates were most uncertain for large waterbodies, suggesting that efforts to improve the accuracy of the lentic N~2~O flux estimate should include additional measurements at the nation’s largest lakes and reservoirs.  Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS waterbodies are a net source of sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source [@epa_2023].
